The Habitable Zone Inner Edge Discontinuity: Probing the Imprint of Runaway Greenhouse Climates on Exoplanet Demographics

### 2023-04-15 src/tex/ms.tex
draft a title
sort out TODOs and minor changes

% Title
\title{The imprint of global magma oceans on exoplanet demographics}

@@

\subsubsection{Atmospheric spectral signatures}\label{sec:dis_atmospheres}
%\todo[inline]{TIM: Ideas?\\
%discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (e.g. Tim's recent talk). Or: high O2 abundance during H loss (Wordsworth \& Pierrehumbert 2013b; Luger \& Barnes 2015, Schaefer et al. 2016)}
\todo[inline]{``Can a smaller sample size's diagnostic value be increased if atmospheric compositional information can be added, i.e., PLATO + ARIEL (or similar combo)?''}
\todo[inline]{also: use atmospheric characterization to identify small planet cores with thick H/He envelopes that contribute to diluting the signal (aka background population). These will be amenable for transmission spec due to large scale heights (Moran2018, \citet{Turbet2019}).\\ Spectral water features  $<<$ rgh-induced radius change~\citep[][end of Sect. 4]{Turbet2019}.}

We currently see four potential lines of discriminating atmospheric signatures associated with super-runaway climates in the exoplanet population.
(i) Detecting the atmospheric windows of water vapor in the near- to mid-infrared.
In a super-runaway state the atmosphere is dominated by the opacity of water vapor, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m~\citep[e.g.][]{Boukrouche2021}.
This would require an instrument with capabilities in this wavelength regime, for instance ELT~METIS, JWST~MIRI, or future missions such as \life\ ~\citep{2019A&A...621A.125B,2022A&A...664A..21Q,2022A&A...664A..22D}. CITE METIS, MIRI!
Potentially the two-band filter capabilities of PLATO may offer insight into particularly pronounced spectral features in this super-runaway regime~\citep{2020SSRv..216...98G}.
(ii) Post-runaway planets may be detectable by O$_{3}$ absorption due to build-up of abiotic oxygen, left-over from photochemical dissociation, and hydrogen loss~\citep{2014ApJ...785L..20W,Luger2015}.
(iii) Water loss in runaway greenhouse planets episodes would change the atmosphere's D/H isotopic ratio, akin to the enhancement in Venus' present-day atmosphere~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
This may be detectable in high-resolution observations focusing on isotope-sensitive transitions~\citep{2019AJ....158...26L,2019A&A...622A.139M}.
(iv) Disequilibrium chemistry in tidally-locked super-runaway planets.
It has been suggested that the atmospheric depth and the presence or absence of oceans on sub-Neptunes could be probed via the abundance of species that are photochemically destroyed in the upper atmosphere, and replenished from either thermochemical layers~\citep{2021ApJ...914...38Y,2021ApJ...922L..27T} or at an ocean-atmosphere interface~\citep{2019ApJ...887..231L,2021ApJ...921L...8H}.
Atmospheric nitrogen and carbon compounds show strong sensitivity in their partitioning coefficients into magma~\citep{2022E&PSL.59817847G,2022PSJ.....3...93B}.
Therefore, it may be possible to discern the presence of an underlying magma ocean if the presence of an atmosphere on a rocky exoplanet can be confirmed.
This has been suggested to be done via (v) eclipse photometry~\citep{2019ApJ...886..141M,2019ApJ...886..140K} through the presence of a high albedo, which is expected to differ from the crystallized rock of a solidified magma ocean~\citep{2020ApJ...898..160E,Fortin2022}.

@@

\section{Conclusions}\label{sec:conclusions}
A robust prediction of runaway greenhouse models is a significant inflation of rocky planet radii.
Using \bioverse, a quantitative hypothesis testing framework, we have explored the potential of contemporary exoplanet missions to statistically detect a radius/density discontinuity resulting from this inflation in the exoplanet population.
Our key findings are as follows:
\begin{enumerate}
    \item The predicted runaway greenhouse transition causes a discontinuity in the radius and density distribution of small exoplanets with respect to the incident host star irradiation.
    \item This habitable zone inner edge discontinuity should be detectable with high-precision transit measurements.
          For a large sample $\gtrsim 100$, a detection is likely if radius inflation occurs on at least \SI{10}{\percent} of the observed planets and if typical water mass fractions are above $\sim 10^{-3}$.
    \item We find that the planned \plato\ transit survey will provide a sufficient sample and the required precision to detect the predicted trend.
          Assuming the projected photometric precision, \plato\ will be able to test the runaway greenhouse hypothesis for planet yields $\gtrsim 100$.
    \item The diagnostic power of transit missions in testing this hypothesis can be increased through a follow-up campaign providing planet mass measurements.
          This can reduce the required planet yield by about a factor of three.
          Only an adequate sample of planets orbiting M~dwarfs will ensure sufficient targets on both sides of the expected threshold instellation.
    \item Testing the runaway greenhouse hypothesis on a population level can provide constraints on the water inventory of rocky exoplanets and thus make an important contribution to assessing their habitability.
         A detection will provide an empirical confirmation of the habitable zone concept and localize its inner edge.

>>>>

% Title
\title{The Habitable Zone Inner Edge Discontinuity: Probing the Imprint of Runaway Greenhouse Climates on Exoplanet Demographics}

@@

\subsubsection{Atmospheric spectral signatures}\label{sec:dis_atmospheres}
%discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (e.g. Tim's recent talk). Or: high O2 abundance during H loss (Wordsworth \& Pierrehumbert 2013b; Luger \& Barnes 2015, Schaefer et al. 2016)}
%\todo[inline]{also: use atmospheric characterization to identify small planet cores with thick H/He envelopes that contribute to diluting the signal (aka background population). These will be amenable for transmission spec due to large scale heights (Moran2018, \citet{Turbet2019}).\\ Spectral water features  $<<$ rgh-induced radius change~\citep[][end of Sect. 4]{Turbet2019}.}

We currently see four potential lines of discriminating atmospheric signatures associated with super-runaway climates in the exoplanet population.
(i) Detecting the atmospheric windows of water vapor in the near- to mid-infrared.
In a super-runaway state the atmosphere is dominated by the opacity of water vapor, which has two prominent spectral features: one at 3.5--4.5 $\mu$m, and one between 8--20 $\mu$m~\citep[e.g.,][]{Boukrouche2021}.
This would require an instrument with capabilities in this wavelength regime, for instance ELT~METIS, JWST~MIRI, or future missions such as \life\ ~\citep{2019A&A...621A.125B,2022A&A...664A..21Q,2022A&A...664A..22D}. CITE METIS, MIRI!
Potentially the two-band filter capabilities of PLATO may offer insight into particularly pronounced spectral features in this super-runaway regime~\citep{2020SSRv..216...98G}.
(ii) Post-runaway planets may be detectable by O$_{3}$ absorption due to build-up of abiotic oxygen, left-over from photochemical dissociation, and hydrogen loss~\citep{2014ApJ...785L..20W,Luger2015}.
(iii) Water loss in runaway greenhouse planets episodes would change the atmosphere's D/H isotopic ratio, akin to the enhancement in Venus' present-day atmosphere~\citep{2019JGRE..124.2015K,2021JGRE..12606643K}.
This may be detectable in high-resolution observations focusing on isotope-sensitive transitions~\citep{2019AJ....158...26L,2019A&A...622A.139M}.
(iv) Disequilibrium chemistry in tidally-locked super-runaway planets.
It has been suggested that the atmospheric depth and the presence or absence of oceans on sub-Neptunes could be probed via the abundance of species that are photochemically destroyed in the upper atmosphere, and replenished from either thermochemical layers~\citep{2021ApJ...914...38Y,2021ApJ...922L..27T} or at an ocean-atmosphere interface~\citep{2019ApJ...887..231L,2021ApJ...921L...8H}.
Atmospheric nitrogen and carbon compounds show strong sensitivity in their partitioning coefficients into magma~\citep{2022E&PSL.59817847G,2022PSJ.....3...93B}.
Therefore, it may be possible to discern the presence of an underlying magma ocean if the presence of an atmosphere on a rocky exoplanet can be confirmed.
This has been suggested to be done via (v) eclipse photometry~\citep{2019ApJ...886..141M,2019ApJ...886..140K} through the presence of a high albedo, which is expected to differ from the crystallized rock of a solidified magma ocean~\citep{2020ApJ...898..160E,Fortin2022}.

@@

\section{Conclusions}\label{sec:conclusions}
Significant inflation of rocky planet radii is a robust prediction of runaway greenhouse models.
Using \bioverse, a quantitative hypothesis testing framework, we have explored the potential of contemporary exoplanet missions to statistically detect a radius/density discontinuity resulting from this inflation in the exoplanet population.
Our key findings are as follows:
\begin{enumerate}
    \item The predicted runaway greenhouse transition causes a discontinuity in the radius and density distribution of small exoplanets with respect to the incident host star irradiation.
    \item This habitable zone inner edge discontinuity should be detectable with high-precision transit measurements.
          For a large sample $\gtrsim 100$, a detection is likely if radius inflation occurs on at least \SI{10}{\percent} of the observed planets and if typical water mass fractions are above $\sim 10^{-3}$.
    \item We find that the planned \plato\ transit survey will provide a sufficient sample and the required precision to detect the predicted trend.
          Assuming the projected photometric precision, \plato\ will be able to test the runaway greenhouse hypothesis for planet yields $\gtrsim 100$.
    \item The diagnostic power of transit missions in testing this hypothesis can be increased through a follow-up campaign providing planet mass measurements.
          This can reduce the required planet yield by about a factor of three.
          Only an adequate sample of planets orbiting M~dwarfs will ensure sufficient targets on both sides of the expected threshold instellation.
    \item Testing the runaway greenhouse hypothesis on a population level can provide constraints on the water inventory of rocky exoplanets and thus make an important contribution to assessing their habitability.
         A detection will provide an empirical confirmation of the habitable zone concept and localize its inner edge.
